The benefits of reporting critical effect size values

### 2024-08-12 paper/paper.Rmd
^^

In the present paper, we suggest that researchers systematically report the critical effect size values when they make inference using statistical significance. The "critical effect size values" refers to the smallest statistically significant effect sizes, which depends on the test performed and the characteristics of the sample [@lakens2022sample]. To cut to the chase, let us consider a study that tests (using for example a two-tailed test) a bivariate correlation with n = 20 participants and uses p < 0.05 as the threshold for statistical significance: in this case, the smallest significant effects (critical values) are Pearson's r = -0.44 and r = 0.44. As we will argue, the critical values can serve as a useful complement to aid in the interpretation of study results, communicate their relevance, help assessing their replicability, particularly when statistical significance is relied upon for inference. This is especially important in cases where the sample size was not (or could not) be predetermined based on a theory-driven target effect size or when statistical power may be difficult to determine or deviates from an optimal level.

Despite longstanding criticisms, the Null Hypothesis Significance Testing (NHST) remains a prominent approach for statistical inference in science [@cohen1994; @gigerenzer2004null; @krueger2001null]. Over the past decades, numerous proposals have emerged to enhance inference based on this approach. These include conducting power analyses to determine sample size (n) a priori based on theory-informed target effects, and reporting of effect sizes and their confidence intervals, accompanied by pertinent related comments, rather than drawing dichotomous conclusions solely based on statistical significance [@transue2019apa]. Even for the most skeptical, there are suggestions that NHST may possess merits of its own. For instance, @wilson2020science propose that it could serve as a useful preliminary filter utilized by "original science" for screening potentially interesting effects, that should subsequently be validated by "replication science".

@@

Let us consider the following two scenarios as examples. First, imagine researchers conducting a study involving a between-group comparison. Due to severe resource constraints, they are only able to collect a limited sample size of n = 30 (n = 15 per group). No statistically significant effect is detected with p < 0.05. Given their prior expectation that the effect of interest may not be large, they acknowledge that their study was likely underpowered, though uncertain to what extent. Subsequently, they compute the critical values, revealing Cohen's d = -0.75 and d = 0.75. This indicates that any estimated |Cohen's d| < 0.75 will certainly fail to reach statistical significance. By reporting these critical values, the researchers transparently convey that estimated effects up to a medium-to-large magnitude will consistently fall short of significance. In essence, this scenario exemplifies a version of the "winner's curse" [@hedges1984estimation], made explicit. In this context, the winner's curse indicates the tendency for a statistically significant initial finding (i.e., a "winner") to be associated with an overestimated effect size which will probably be deflated in subsequent replications (i.e., the "curse"), especially if such initial finding was obtained on a small sample. Knowing from the critical values that a specific result must necessarily be associated with an estimated large effect size or be non-significant, readers are, in principle, prevented from both over-interpreting the magnitude of an effect size and from equating lack of statistical significance with lack of an effect of interest.

In a second scenario, imagine researchers who gain access to a very large archival dataset (n = 5,000) and decide to explore some bivariate correlations. With such a large sample size, researchers are not constrained by sample size limitations and could instead just focus on effect sizes. However, they opt to legitimately use statistical significance of a two-tailed test as their primary criterion for whether to comment on effects in their further discussions. In this context, the critical effect size values for significance are determined to be correlations of $r = \pm 0.03$. This implies that practically any effect differing from zero, regardless of its small magnitude, may easily attain statistical significance. It is worth noting that such small effects may potentially reflect just minor artifacts [@wilson2020science], such as subtle experimenter effects or slight non-independence among observations, even in meticulously designed studies. While readers may be tempted to interpret any reported effects simply because they attain statistical significance, preliminarily signaling the critical value level can serve as a clear warning that the significance filter might be somehow permissive, thus urging interpretive caution.

@@

To provide more tangible illustrations of the practical application of critical values, let us examine two real-world instances drawn from published research. The first one features a small sample, and has been targeted in the series of replications by the Open Science Collaboration [@open2015estimating], while the second one presents with a very large sample and has already attracted attention in the scientific community. The first case pertains to "Study 5" conducted by @mccrea2008self. This experiment compared two groups based on their performance percentage in a math test following a preliminary practice session with feedback of failure. One group was exposed to failure-excusing, self-handicapping thoughts, while both groups underwent the math test afterward. With a modest sample size of 28 participants split into two groups (13 and 15 participants, respectively), a directional one-tailed t-test was conducted on the percentage of correct answers to compare the groups. The resulting effect was statistically significant, $t(26) = 1.87, p (one-tailed) < .05$. Based on the reported data, Cohen's d = 0.736, indicating a considerable effect size. However, the critical value for Cohen's d in this context was 0.645, suggesting that, given this sample size, statistical significance is achieved only for estimated effects of substantial magnitude. Subsequent replication efforts [@open2015estimating] reaffirmed the significance of the effect, and used a somewhat larger sample (n = 61), $t(59)=2.325$. The estimated effect size was somehow lower than that originally reported, however, $d = 0.605$ (here, critical d = 0.428). The second scenario involves the study by @kramer2014experimental, which explored the impact of emotional content on Facebook users' experiences. With a notably large sample size of n = 689'003, the researchers observed a significant effect worsened emotional states when positive posts were reduced, $t(310,044) = −5.63$, $p < 0.001$ (non-directional test). Based on this and other results of a similar, the authors conclude that emotions expressed by other users on Facebook influence our own emotions. Despite the statistical significance, however, the critical values for Cohen's d in this case are -0.006 and +0.006, emphasizing the importance of considering effect size beyond mere significance. This prompts reflection on whether an effect of such magnitude on such a large sample (actual d = 0.02) is truly meaningful in individuals' lives.

@@

In this section, we provide guidance and formulas for computing critical values in general, with examples for frequently encountered effect sizes including Standardized Mean Differences (Cohen's d), correlations (Pearson's r), and raw and standardized coefficients in linear models. These formulas have been incorporated into R functions of the package "criticalESvalues" accessible at: [xxxxx], and are elaborated upon in the subsequent section.

@@

The output now gives a wider range of values: the cohen's d calculated on the data (d), the critical cohen's d (dc), the numerator of the formula for the critical cohen's d (bc), the cohen's d adjusted for small samples (g) and the critical cohen's d adjusted for small samples (gc).

@@

The direction of the hypothesis and the test to apply, either t-test or z-test, should be specified. The output will return the critical correlation value, the degrees of freedom and the type of test used.

@@

An advantage of reporting the critical value is that it can be precisely computed in any scenario, without requiring assumptions about the expected effect size, as is the case with power calculations. The critical value represents a directly interpretable benchmark that is especially useful in situations where statistical power is below the desired level and researchers are left otherwise uncertain about how to proceed with the interpretation of a study findings. For example, let us say that we read a published article reporting some effects as statistically significant, while others as not: we suspect that the study may be underpowered, but we are widely uncertain about the magnitude of possible true effects. To what extent can the reported results be interpreted, precisely? Knowing the critical value provides us with a clear benchmark. Conversely, let us say that an effect achieves significance in a very large sample: researchers tend to draw substantive conclusions based on this. But is it of real theoretical relevance? If in comparing two groups, such as controls versus treatment, any Cohen's d > 0.07 would reach significance, is statistical significance enough to signal a "successful" treatment? Maybe yes, even if effects are tiny [e.g., @funder2019evaluating], but knowing the critical value certainly prompts some appropriate interpretive caution.

Reporting the critical value(s) can also be an efficient way to allow researchers to evaluate which findings are statistically significant. For example, in a correlation table researchers customarily add an asterisk to all statistically significant correlations. But as long as all correlations are based on the same sample size, researchers can simply remark 'the critical effect size is *r* = 0.3' and readers will know all correlations larger than this value are statistically significant.

@@

Real-case scenarios may not always be that simple. Hence, we chose to expand the application of computing critical significance values beyond Cohen's d and correlation to include linear regression with both raw and standardized coefficients. This serves as a first step in computing critical values for a wider array of effects encountered in practical scenarios, where linear models and their extensions are commonly utilized for modeling purposes. A prerequisite is that researchers must be able to identify what parameters in their statistical models reflect the effect sizes of interest, and that they are able to assess their relevance. Notably, however, this prerequisite aligns with the requirements of APA style guidelines concerning the reporting of effect sizes. For further illustration and application, additional examples are provided in the Supplementary online material.

>>>>

In the present paper, we suggest that researchers systematically report the critical effect size values when they make inference using statistical significance. The "critical effect size values" refers to the smallest statistically significant effect sizes, which depends on the test performed and the characteristics of the sample [@lakens2022sample]. To cut to the chase, let us consider a study that tests (using for example a two-tailed test) a bivariate correlation with $n = 20$ participants and uses $p < 0.05$ as the threshold for statistical significance: in this case, the smallest significant effects (critical values) are Pearson's $r = -0.44$ and $r = 0.44$. As we will argue, the critical values can serve as a useful complement to aid in the interpretation of study results, communicate their relevance, help assessing their replicability, particularly when statistical significance is relied upon for inference. This is especially important in cases where the sample size was not (or could not) be predetermined based on a theory-driven target effect size or when statistical power may be difficult to determine or deviates from an optimal level.

Despite longstanding criticisms, the Null Hypothesis Significance Testing (NHST) remains a prominent approach for statistical inference in science [@cohen1994; @gigerenzer2004null; @krueger2001null]. Over the past decades, numerous proposals have emerged to enhance inference based on this approach. These include conducting power analyses to determine sample size ($n$) a priori based on theory-informed target effects, and reporting of effect sizes and their confidence intervals, accompanied by pertinent related comments, rather than drawing dichotomous conclusions solely based on statistical significance [@transue2019apa]. Even for the most skeptical, there are suggestions that NHST may possess merits of its own. For instance, @wilson2020science propose that it could serve as a useful preliminary filter utilized by "original science" for screening potentially interesting effects, that should subsequently be validated by "replication science".

@@

Let us consider the following two scenarios as examples. First, imagine researchers conducting a study involving a between-group comparison. Due to severe resource constraints, they are only able to collect a limited sample size of $n = 30$ ($n = 15$ per group). No statistically significant effect is detected with $p < 0.05$. Given their prior expectation that the effect of interest may not be large, they acknowledge that their study was likely underpowered, though uncertain to what extent. Subsequently, they compute the critical values, revealing Cohen's $d = -0.75$ and $d = 0.75$. This indicates that any estimated |Cohen's $d| < 0.75$ will certainly fail to reach statistical significance. By reporting these critical values, the researchers transparently convey that estimated effects up to a medium-to-large magnitude will consistently fall short of significance. In essence, this scenario exemplifies a version of the "winner's curse" [@hedges1984estimation], made explicit. In this context, the winner's curse indicates the tendency for a statistically significant initial finding (i.e., a "winner") to be associated with an overestimated effect size which will probably be deflated in subsequent replications (i.e., the "curse"), especially if such initial finding was obtained on a small sample. Knowing from the critical values that a specific result must necessarily be associated with an estimated large effect size or be non-significant, readers are, in principle, prevented from both over-interpreting the magnitude of an effect size and from equating lack of statistical significance with lack of an effect of interest.

In a second scenario, imagine researchers who gain access to a very large archival dataset ($n = 5,000$) and decide to explore some bivariate correlations. With such a large sample size, researchers are not constrained by sample size limitations and could instead just focus on effect sizes. However, they opt to legitimately use statistical significance of a two-tailed test as their primary criterion for whether to comment on effects in their further discussions. In this context, the critical effect size values for significance are determined to be correlations of $r = \pm 0.03$. This implies that practically any effect differing from zero, regardless of its small magnitude, may easily attain statistical significance. It is worth noting that such small effects may potentially reflect just minor artifacts [@wilson2020science], such as subtle experimenter effects or slight non-independence among observations, even in meticulously designed studies. While readers may be tempted to interpret any reported effects simply because they attain statistical significance, preliminarily signaling the critical value level can serve as a clear warning that the significance filter might be somehow permissive, thus urging interpretive caution.

@@

To provide more tangible illustrations of the practical application of critical values, let us examine two real-world instances drawn from published research. The first one features a small sample, and has been targeted in the series of replications by the Open Science Collaboration [@open2015estimating], while the second one presents with a very large sample and has already attracted attention in the scientific community. The first case pertains to "Study 5" conducted by @mccrea2008self. This experiment compared two groups based on their performance percentage in a math test following a preliminary practice session with feedback of failure. One group was exposed to failure-excusing, self-handicapping thoughts, while both groups underwent the math test afterward. With a modest sample size of 28 participants split into two groups (13 and 15 participants, respectively), a directional one-tailed t-test was conducted on the percentage of correct answers to compare the groups. The resulting effect was statistically significant, $t(26) = 1.87, p (one-tailed) < .05$. Based on the reported data, Cohen's $d = 0.736$, indicating a considerable effect size. However, the critical value for Cohen's $d$ in this context was $0.645$, suggesting that, given this sample size, statistical significance is achieved only for estimated effects of substantial magnitude. Subsequent replication efforts [@open2015estimating] reaffirmed the significance of the effect, and used a somewhat larger sample ($n = 61$), $t(59)=2.325$. The estimated effect size was somehow lower than that originally reported, however, $d = 0.605$ (here, critical d = 0.428). The second scenario involves the study by @kramer2014experimental, which explored the impact of emotional content on Facebook users' experiences. With a notably large sample size of $n = 689'003$, the researchers observed a significant effect worsened emotional states when positive posts were reduced, $t(310,044) = −5.63$, $p < 0.001$ (non-directional test). Based on this and other results of a similar, the authors conclude that emotions expressed by other users on Facebook influence our own emotions. Despite the statistical significance, however, the critical values for Cohen's $d$ in this case are $-0.006$ and $+0.006$, emphasizing the importance of considering effect size beyond mere significance. This prompts reflection on whether an effect of such magnitude on such a large sample (actual $d = 0.02$) is truly meaningful in individuals' lives.

@@

In this section, we provide guidance and formulas for computing critical values in general, with examples for frequently encountered effect sizes including Standardized Mean Differences (Cohen's $d$), correlations (Pearson's r), and raw and standardized coefficients in linear models. These formulas have been incorporated into R functions of the package "criticalESvalues" accessible at: [xxxxx], and are elaborated upon in the subsequent section.

@@

The output now gives a wider range of values: the Cohen's $d$ calculated on the data ($d$), the critical Cohen's $d$ ($dc$), the numerator of the formula for the critical Cohen's $d$ ($bc$), the Cohen's $d$ adjusted for small samples ($g$) and the critical Cohen's $d$ adjusted for small samples ($gc$).

@@

The direction of the hypothesis and the test to apply, either $t$-test or $z$-test, should be specified. The output will return the critical correlation value, the degrees of freedom and the type of test used.

@@

An advantage of reporting the critical value is that it can be precisely computed in any scenario, without requiring assumptions about the expected effect size, as is the case with power calculations. The critical value represents a directly interpretable benchmark that is especially useful in situations where statistical power is below the desired level and researchers are left otherwise uncertain about how to proceed with the interpretation of a study findings. For example, let us say that we read a published article reporting some effects as statistically significant, while others as not: we suspect that the study may be underpowered, but we are widely uncertain about the magnitude of possible true effects. To what extent can the reported results be interpreted, precisely? Knowing the critical value provides us with a clear benchmark. Conversely, let us say that an effect achieves significance in a very large sample: researchers tend to draw substantive conclusions based on this. But is it of real theoretical relevance? If in comparing two groups, such as controls versus treatment, any Cohen's $d > 0.07$ would reach significance, is statistical significance enough to signal a "successful" treatment? Maybe yes, even if effects are tiny [e.g., @funder2019evaluating], but knowing the critical value certainly prompts some appropriate interpretive caution.

Reporting the critical value(s) can also be an efficient way to allow researchers to evaluate which findings are statistically significant. For example, in a correlation table researchers customarily add an asterisk to all statistically significant correlations. But as long as all correlations are based on the same sample size, researchers can simply remark 'the critical effect size is $r = 0.3$' and readers will know all correlations larger than this value are statistically significant.

@@

Real-case scenarios may not always be that simple. Hence, we chose to expand the application of computing critical significance values beyond Cohen's $d$ and correlation to include linear regression with both raw and standardized coefficients. This serves as a first step in computing critical values for a wider array of effects encountered in practical scenarios, where linear models and their extensions are commonly utilized for modeling purposes. A prerequisite is that researchers must be able to identify what parameters in their statistical models reflect the effect sizes of interest, and that they are able to assess their relevance. Notably, however, this prerequisite aligns with the requirements of APA style guidelines concerning the reporting of effect sizes. For further illustration and application, additional examples are provided in the Supplementary online material.
